Assessment of zonally symmetric and asymmetric components of the Southern Annular Mode using a novel approach

### 2021-04-28 analysis/paper/asymsam.Rmd
Cambia signo en Figura A6 y agrega que hay cambios.
#23

By performing the EOF analysis using data for all months we are assuming that the zonally asymmetric structure of the SAM is the same at all seasons.
The latter was assessed by computing geopotential height zonal anomalies by projecting the first EOF of each season independently.
The following seasons were considered -- December to February (DJF), March to May (MAM), June to August (JJA) and September to November (SON).
Results are very similar to each other in the troposphere (Figure \@ref(fig:A5), row 2) and show spatial correlations between 0.65 (DJF with JJA) and 0.9 (MAM with SON).
In the stratosphere (Figure \@ref(fig:A5), row 1), patterns are similar for all seasons except DJF, when the wave-1 zonal anomalies are rotated 90\degree in comparison with the rest of the year.
Spatial correlations in the stratosphere are between -0.24 (DJF with SON) and 0.95 (MAM with JJA).
Therefore, the results confirm that the zonal asymmetric structure of the SAM is very similar throughout most of the year.
The only outlier being DJF, which shows much lower correlations with the other seasons at both levels and the weakest zonal anomalies (Figure \@ref{fig:A3}), which is consistent with @fogt2020.
With this in mind, a

The method also assumes that the zonally asymmetric pattern of the SAM remains stationary along the period considered.
@silvestri2009 suggest that this might not be the case between 1958 and 2004.
Zonal asymmetric patterns of SAM were computed for the two halves of the period (1979 to 1998 and 1999 to 2018) respectively.
They show no systematic change neither in the stratosphere nor in the troposphere (Figure \@ref(fig:A6)).

@@

```{r A6, fig.cap = "Regression of 50~hPa and 700~hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for the periods 1979 -- 1998 and 1999 -- 2018. Spatial correlation between both fields is 0.86 for the 50~hPa fields and 0.76 for the 700~hPa fields.", fig.width=6, fig.height=5}
set.seed(42)
SAM_prepost <- hgt[lat <= -20] %>% 
  .[lev %in% c(700, 50)] %>%
  .[, hgt := hgt_a*sqrt(cos(lat*pi/180))] %>% 
  .[, year := year(time)] %>% 
  .[, period := factor(ifelse(year(time) <= 1998, "pre", "post"),
                       levels = c("pre", "post"), ordered = TRUE,
                       labels = c("1979-1998", "1999-2018"))] %>%
  .[, .(eof = list(EOF(hgt ~ time | lon + lat, n = 1, data = .SD))), 
    by = .(period, lev)]

@@

patterns %>% 
  periodic(lon = c(0, 360)) %>%
  ggplot(aes(lon, lat)) +
  geom_contour_fill(aes(z = estimate_a), breaks = AnchorBreaks(0, 5, 0)) +
  geom_contour_tanaka(aes(z = estimate_a), breaks = AnchorBreaks(0, 5, 0)) +
  geom_qmap() +
  scale_x_longitude() +
  scale_y_latitude(limits = c(NA, -20), labels = NULL) +
  scale_fill_divergent(NULL, 
                       guide = guide_colorstrip_bottom(),
                       breaks = AnchorBreaks(0, 5, exclude = 0)) +
  coord_polar() +
  facet_grid(lev~period, labeller = labeller(lev = lev.lab)) +
  axis_labs_smol +
  no_grid 
```

>>>>

By performing the EOF analysis using data for all months we are assuming that the zonally asymmetric structure of the SAM is the same at all seasons.
The latter was assessed by computing geopotential height zonal anomalies by projecting the first EOF of each season independently.
The following seasons were considered -- December to February (DJF), March to May (MAM), June to August (JJA) and September to November (SON).
Results are very similar to each other in the troposphere (Figure \@ref(fig:A5), row 2) and show spatial correlations between 0.65 (DJF with JJA) and 0.9 (MAM with SON).
In the stratosphere (Figure \@ref(fig:A5), row 1), patterns are similar for all seasons except DJF, when the wave-1 zonal anomalies are rotated 90\degree in comparison with the rest of the year.
Spatial correlations in the stratosphere are between -0.24 (DJF with SON) and 0.95 (MAM with JJA).
Therefore, the results confirm that the zonal asymmetric structure of the SAM is very similar throughout most of the year.
The only outlier being DJF, which shows much lower correlations with the other seasons at both levels and the weakest zonal anomalies (Figure \@ref{fig:A3}), which is consistent with @fogt2020.

The method also assumes that the zonally asymmetric pattern of the SAM remains stationary along the period considered.
@silvestri2009 suggest that this might not be the case between 1958 and 2004.
Zonal asymmetric patterns of SAM were computed for the two halves of the period (1979 to 1998 and 1999 to 2018) respectively.
The differences between the two periods appear to be relatively small in both the troposphere and stratosphere (Figure \@ref(fig:A6)).

@@

```{r A6, fig.cap = "Regression of 50~hPa and 700~hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for the periods 1979 -- 1998 and 1999 -- 2018. Spatial correlation between both fields is 0.86 for the 50~hPa fields and 0.76 for the 700~hPa fields.", fig.width=6, fig.height=5}
set.seed(42)
SAM_prepost <- hgt[lat <= -20] %>% 
  .[lev %in% c(700, 50)] %>%
  .[, hgt := hgt_a*sqrt(cos(lat*pi/180))] %>%

  .[, year := year(time)] %>% 
  .[, period := factor(ifelse(year(time) <= 1998, "pre", "post"),
                       levels = c("pre", "post"), ordered = TRUE,
                       labels = c("1979-1998", "1999-2018"))] %>%
  .[, .(eof = list(EOF(hgt ~ time | lon + lat, n = 1, data = .SD))), 
    by = .(period, lev)]

@@

patterns %>% 
  .[lev == 50 & period == "1999-2018", estimate_a := -estimate_a] %>% 
  periodic(lon = c(0, 360)) %>%
  ggplot(aes(lon, lat)) +
  geom_contour_fill(aes(z = estimate_a), breaks = AnchorBreaks(0, 5, 0)) +
  geom_contour_tanaka(aes(z = estimate_a), breaks = AnchorBreaks(0, 5, 0)) +
  geom_qmap() +
  scale_x_longitude() +
  scale_y_latitude(limits = c(NA, -20), labels = NULL) +
  scale_fill_divergent(NULL, 
                       guide = guide_colorstrip_bottom(),
                       breaks = AnchorBreaks(0, 5, exclude = 0)) +
  coord_polar() +
  facet_grid(lev~period, labeller = labeller(lev = lev.lab)) +
  axis_labs_smol +
  no_grid 
```
